---
title: "Prosper Loan Data Analysis"
author: "Alexander Baranov"
date: "30 August 2015"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      fig.path='Figs/',
                      echo=FALSE, warning=FALSE,
                      message=FALSE)
```

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
library(RColorBrewer)
```

The dataset is about prosper loans. It contains records for number of person to person loans. It would be interesting to check who borrow money, what amount, what rate and how rate depends on other attributes.

# Exploration of the data
```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
loanData <- read.csv("prosperLoanData.csv")
```

Add ListingCategory factor variable that meets int ListingCategory..numeric. column

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
loanData$ListingCategory <-as.factor(loanData$ListingCategory..numeric.)
levels(loanData$ListingCategory) <- c("Not Available",
                                      "Debt Consolidation",
                                      "Home Improvement",
                                      "Business",
                                      "Personal Loan",
                                      "Student Use",
                                      "Auto",
                                      "Other",
                                      "Baby&Adoption",
                                      "Boat",
                                      "Cosmetic Procedure",
                                      "Engagement Ring",
                                      "Green Loans",
                                      "Household Expenses",
                                      "Large Purchases",
                                      "Medical/Dental",
                                      "Motorcycle",
                                      "RV",
                                      "Taxes",
                                      "Vacation",
                                      "Wedding Loans")
```


## Univariate Plots Section

Check distributions for some basic attributes.

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
names(loanData)
summary(loanData)
```

Unfortunately I was unable to run gpairs on my laptop. It worked too long anf inally hung each time. I did subsetting of several columns and run ggpairs against narrowed dataset.
Finally I skip it from this report due its large size and less informativity.

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
loanData.forggpairs <- loanData %>%
  select(ListingCategory,
         Term,
         #BorrowerAPR,
         BorrowerRate,
         LenderYield,
         EstimatedEffectiveYield,
         #EstimatedLoss,
         #EstimatedReturn,
         ProsperScore,
         #Occupation,
         EmploymentStatus,
         #EmploymentStatusDuration,
         #IsBorrowerHomeowner,
         #CreditScoreRangeLower,
         #CreditScoreRangeUpper,
         CurrentCreditLines,
         #OpenCreditLines,
         #TotalCreditLinespast7years,
         #OpenRevolvingAccounts,
         #OpenRevolvingMonthlyPayment,
         #InquiriesLast6Months,
         #TotalInquiries,
         #CurrentDelinquencies,
         AmountDelinquent,
         #DelinquenciesLast7Years,
         #RevolvingCreditBalance,
         #BankcardUtilization,
         #AvailableBankcardCredit,
         DebtToIncomeRatio,
         IncomeRange)
         #IncomeVerifiable,
         #StatedMonthlyIncome)
# ggpairs(loanData.forggpairs)
```

Let's start with CreditGrade

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = CreditGrade, data = loanData)
```

Urm.. Most of records without CreditGrade? It looks like it is due 'Applicable for listings pre-2009 period and will only be populated for those listings.'
Try without NA data

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = CreditGrade, 
      data = subset(loanData, !is.na(CreditGrade) &
                      CreditGrade != ''))
```


Ok. Further... Term in months

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = Term, data = loanData, binwidth = 6)
```

It looks like there are 3 standard terms. Check terms in years

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = Term/12, data = loanData, binwidth = 1)
```

As it could be expected - 1, 3, and 5 years.

Categories of the listing: 0 - Not Available, 1 - Debt Consolidation, 2 - Home Improvement, 3 - Business, 4 - Personal Loan, 5 - Student Use, 6 - Auto, 7- Other, 8 - Baby&Adoption, 9 - Boat, 10 - Cosmetic Procedure, 11 - Engagement Ring, 12 - Green Loans, 13 - Household Expenses, 14 - Large Purchases, 15 - Medical/Dental, 16 - Motorcycle, 17 - RV, 18 - Taxes, 19 - Vacation, 20 - Wedding Loans

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = ListingCategory, data = loanData, binwidth = 1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

So top-3 reasons excluding N/A are following
1. Debt Consolidation
2. Other
3. Home Improvement

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Check EmploymentStatus without records for which EmploymentStatus is unknown
qplot(x = EmploymentStatus, 
      data = subset(loanData, !is.na(EmploymentStatus) &
                      EmploymentStatus != 'Not available' &
                      EmploymentStatus != ''))
```

Not surprisingly that most of loans are for employed and full-time employed. Lender needs to be sure that loan will be payed. Most probability of returns for people who has costant income. Such people is employed.

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = BorrowerRate, data = loanData)
```

Explore it per possible groups. Starts with BorrowerRate per Term

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = BorrowerRate, data = loanData) +
  facet_wrap(~Term)
```

Continue with BorrowerRate per IncomeRange

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = BorrowerRate, data = loanData) +
  facet_wrap(~IncomeRange)
```

Check how DebtToIncomeRatio is distributed

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = DebtToIncomeRatio, 
      data = subset(loanData, 
                    DebtToIncomeRatio <=
                      quantile(DebtToIncomeRatio, 0.99,
                               na.rm = TRUE)))
```

Looks like Poisson distribution.
Check how LoanOriginalAmount is distributed

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = LoanOriginalAmount, data = loanData)
```

Peaks on 500, 1000, 1500, 2000, amd 2500. Returning back to Listing histogram we saw that most popular categories are debt consolidation and other. People tend to borrow round amount of money like 500, 1000, 2000 etc
Further check MonthlyLoanPayment

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = MonthlyLoanPayment, data = loanData)
```

and IncomeRange

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
qplot(x = IncomeRange, 
      data = subset(loanData, !is.na(IncomeRange) & 
                      IncomeRange != 'Not employed' & 
                      IncomeRange != 'Not displayed'))

```

## Univariate Analysis

### What is the structure of your dataset?

There are more than 110 thousands observations with 81 variables.

### What is/are the main feature(s) of interest in your dataset?

Most interesting features I believe are BorrowerRate and LoanOriginalAmount. I am going to explore how other attributes affect them.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

For the first glance they are:
- Term
- ListingCategory
- EmploymentStatus
- DebtToIncomeRatio
- IncomeRange

### Did you create any new variables from existing variables in the dataset?

Not yet

## Bivariate Plots Section

Start with checking BorrowerRate per ListingCategory

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
ggplot(aes(x = ListingCategory, y = BorrowerRate), 
       data = loanData) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Rate differs slightly on purpose of loan. Most expensive rates for Cosmetic Procedures and Houshold Expenses. Those purposes are considere too risky and so rate is high. Cheapest loan is Personal Loan. I believe it is due lender and borrower know each other well enough to be sure in borrowers ability to return. 
Continue with LoanOriginalAmount per ListingCategory

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
ggplot(aes(x = ListingCategory, y = LoanOriginalAmount), 
       data = loanData) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Check how BorrowerRate changes with LoanOriginalAmount

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
#loanData.by_ListingCategory <- loanData %>%
#  group_by(ListingCategory) %>%
#  summarize(LoanOriginalAmount_mean = #mean(LoanOriginalAmount),
#            BorrowerRate_mean = mean(BorrowerRate),
#            n = n())
ggplot(aes(x = LoanOriginalAmount, y = BorrowerRate), 
       data = loanData) +
  geom_point(alpha = 1/20) +
  geom_smooth()
```

Consider dataset from another point. Per EmploymentStatus groups. Start with BorrowerRate

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# consider only records for which EmploymentStatus is known
loanData.employed <- subset(loanData, !is.na(EmploymentStatus) &
                             EmploymentStatus != 'Not available' &
                             EmploymentStatus != '')
ggplot(aes(x = EmploymentStatus, y = BorrowerRate), 
       data = loanData.employed) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Continue with LoanOriginalAmount

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
ggplot(aes(x = EmploymentStatus, y = LoanOriginalAmount), 
       data = loanData.employed) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Now check BorrowerRate from DebtToIncomeRatio.

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Drop top 1 % and NA records
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerRate), 
       data = subset(loanData, 
                     DebtToIncomeRatio <= quantile(DebtToIncomeRatio, 
                                                   0.99, na.rm = TRUE))) +
  geom_point(alpha = 1/100) + 
  geom_line(stat = 'summary', fun.y = mean, linetype = 2, color = 2)
```

Urm... It looks like we need smoothing

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Drop top 1 % and NA records
# Add smootinhg
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerRate), 
       data = subset(loanData, 
                     DebtToIncomeRatio <= quantile(DebtToIncomeRatio, 
                                                   0.99, na.rm = TRUE))) +
  geom_point(alpha = 1/100) +
  geom_line(stat = 'summary', fun.y = mean, linetype = 2, color = 2) +
  geom_smooth()
```

Remove original mean line smoothing

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Drop top 1 % and NA records
# Add smootinhg, remove mean line without smoothing
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerRate), 
       data = subset(loanData, 
                     DebtToIncomeRatio <= quantile(DebtToIncomeRatio, 
                                                   0.99, na.rm = TRUE))) +
  geom_point(alpha = 1/100) +
  geom_smooth()
```

Now it is obvious that rate is growing with growth of DebtToIncomeRatio.

Now check a few more dependencies to see if we can find anything interesting.

LoanOriginalAmount ~ IncomeRange

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Drop NA, Not employed and Not displayed records
ggplot(aes(x = IncomeRange, y = LoanOriginalAmount), 
       data = subset(loanData,
                     !is.na(IncomeRange) &
                       IncomeRange != 'Not employed' &
                       IncomeRange != 'Not displayed')) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

BorrowerRate ~ IncomeRange

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Drop NA, Not employed and Not displayed records
ggplot(aes(x = IncomeRange, y = BorrowerRate), 
       data = subset(loanData,
                     !is.na(IncomeRange) &
                       IncomeRange != 'Not employed' &
                       IncomeRange != 'Not displayed')) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

LoanOriginalAmount ~ Term

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
ggplot(aes(x = jitter(Term), y = LoanOriginalAmount), data = loanData) +
  geom_point(alpha = 1/50) +
  geom_smooth()
```

LoanOriginalAmount grows with Term

BorrowerRate ~ Term

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
ggplot(aes(x = jitter(Term), y = BorrowerRate), data = loanData) +
  geom_point(alpha = 1/50) +
  geom_smooth(color=2)
```

BorrowerRate gorws with Term from 1 till 3 years and further remains about constant

## Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

As we see from picture of BorrowRate ~ DebtToIncomeRatio and BorrowRate ~ Term the rate depends at least on DebtToIncomeRatio and Term (there are possible other paramaeters affecting it). A bit unexpectedly to me that it does not depend on IncomeRange. BorrowerRate grows with Term from 1 till 3 years and further remains about constant

## Multivariate Plots Section

I'm providing graph advised by tutor.

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Advised by tutor
# Drop top 1 % and NA records
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerRate,
           color = IncomeRange), 
       data = subset(loanData,
                     IncomeRange != 'Not displayed' &
                     DebtToIncomeRatio <=                                           quantile(DebtToIncomeRatio,
                                  0.99,
                                  na.rm = TRUE))) +
  geom_point(alpha=1/2) + geom_smooth(size=1) +
  scale_colour_brewer(palette = 1)
```

For most of groups BorrowRate grows with DebtToIncomeRatio. Point that meets 0 is higher with lower IncomeRange.
It is interesting curve for Not employed group. It is not monotone. I think it is due other attributes affects much BorrowRate for this group.

# Final Plots and Summary

## Plot One

This plot shows dependency of rate on debt to income ratio. Indeed, people who have low debt in comparision to income are expected to return the loan most probably. On the other hand people having high debt load on their budget are close to default and so lending them money is more risky which should make the rate higher. The plot shows it. Also there is a stable part of rate curve (between abot 5 and 25 %). People with ration in that interval are considered as normal borrowers.

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Drop top 1 % and NA records
# Add smootinhg
ggplot(aes(x = DebtToIncomeRatio * 100, y = BorrowerRate * 100), 
       data = subset(loanData, 
                     DebtToIncomeRatio <= quantile(DebtToIncomeRatio,
                                                   0.99, na.rm = TRUE))) +
  geom_point(alpha = 1/100) +
#  geom_line(stat = 'summary', fun.y = mean, linetype = 2, color = 2) +
  geom_smooth() +
  coord_cartesian(ylim=c(0, 40)) +
  xlab("Debt To Income Ratio, %") +
  ylab("Borrower Rate, %") +
  ggtitle("BorrowRate from DebtToIncomeRatio")
```

## Plot Two

Another attribute Borrower Rate depends on is term of loan. The rate is growing from 12 to 36 month and further it is about stable up to 60 (which is max term)

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
ggplot(aes(x = jitter(Term), y = BorrowerRate * 100), data = loanData) +
  geom_point(alpha = 1/50) +
  geom_smooth(color=2) +
  xlab("Term, months") +
  ylab("Borrower Rate, %") +
  ggtitle("BorrowRate from Term in months")
```

## Plot Three

It is like Plot One but with more details per group of Income Range.
In addition to findings of Plot One there are following ideas:
- People with lowest income range have higher rate
- People that are not employed do not have momotone curve. It is possible due other attributes are more influencing for such group
- Different groups have different max DebtToIncomeRatio. Lowest one is for Not employed people

```{r, echo=FALSE, cache=TRUE, cache.path = 'cache/', fig.path='figure/'}
# Drop top 1 % and NA records
ggplot(aes(x = DebtToIncomeRatio * 100, y = BorrowerRate * 100,
           color = IncomeRange), 
       data = subset(loanData,
                     IncomeRange != 'Not displayed' &
                     DebtToIncomeRatio <=                                           quantile(DebtToIncomeRatio,
                                  0.99,
                                  na.rm = TRUE))) +
  geom_point(alpha=1/2) + geom_smooth(size=1) +
  scale_colour_brewer(palette = 1) +
  xlab("Debt To Income Ratio, %") +
  ylab("Borrower Rate, %") +
  ggtitle("BorrowRate from DebtToIncomeRatio per IncomeRange groups")
```

# Reflection
Data set contains many attributes. I selected BorrowerRate, DebtToIncomeRatio and LoanOriginalAmount as most interesting attributes.
I had technical issues with selecting attributes. ggpair worked too mahc and hung on the whole set so I had to manually filter part of attributes. To make it propriately I had to dig into business side. I understood part of attributes from my previous experience but about half of them were not clear.
During exploration some times I proved my expectations regarding dependencies, sometimes I did not. I expected that rate depends on DebtToIncomeRatio. Though I would also expect that it depends on borrower income though it seems does not.
If I had more time I would improve Multivariate Plots Section and make composite attributes. It'd be interesting to check how not emplyed people are scored for loan, since they seem not have stable income and so they should be scored on differet way than usual.